---
title: "Retail Transactions:  Monthly Report"
author: "Martin Ng"
output: html_document
params:
  views_file: NULL
  transactions_file: NULL
  products_file: NULL
  customers_file: NULL
  month: NULL
  year: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r seed}
set.seed(41)
```

```{r libraries}
library(data.table)
library(DT)
```

```{r constants}

```

```{r functions}
### Q1
# Function to get the top 10 most viewed products for the first 7 days of any given month
get_top_10_views <- function(views_data, products_data, year, month) {
  
  # Step 1: Construct the date range for the first 7 days of the specified month and year
  start_date <- as.POSIXct(paste0(year, "-", sprintf("%02d", month), "-01 00:00:00"))
  end_date <- as.POSIXct(paste0(year, "-", sprintf("%02d", month), "-08 00:00:00"))
  
  # Step 2: Filter the views data for the first 7 days of the specified month and year
  views_first7days <- views_data[time >= start_date & time < end_date]
  
  # Step 3: Count the number of views per product_id
  view_counts <- views_first7days[, .(view_count = .N), by = product_id]
  
  # Step 4: Join with products data to include category
  top_10_views <- merge(view_counts, products_data, by = "product_id", all.x = TRUE)
  
  # Step 5: Reorder the columns to product_id, category, and view_count
  top_10_views <- top_10_views[, .(product_id, category, view_count)]
  
  # Step 6: Sort by view count and get the top 10 most viewed products
  top_10_views <- top_10_views[order(-view_count)][1:10]
  
  # Return the top 10 most viewed products
  return(top_10_views)
}

### Q2
# Function to get the top 10 most viewed products for each category
get_top_10_by_category <- function(views_data, products_data) {

  # Step 1: Count the number of views per product_id
  view_counts <- views_data[, .(view_count = .N), by = product_id]
  
  # Step 2: Join with products data to include category
  view_counts_with_category <- merge(view_counts, products_data, by = "product_id", all.x = TRUE)
  
  # Step 3: For each category, find the top 10 products
  categories <- unique(view_counts_with_category$category)
  results <- list()
  
  for (cat in categories) {
    top_10 <- view_counts_with_category[category == cat][order(-view_count)][1:10, .(product_id, view_count)]
    results[[cat]] <- top_10
  }
  return(results)
}

### Q3
# Function to calculate revenue by category
calculate_revenue_by_category <- function(transactions_data, products_data) {
  # Step 1: Select only the category and product_id from products
  products_reduced <- products_data[, .(product_id, category)]
  
  # Step 2: Merge only the category column into transactions_month
  transactions_with_category <- merge(transactions_data, products_reduced, by = "product_id", all.x = TRUE)
  
  # Step 3: Calculate the total revenue per transaction (price * quantity)
  transactions_with_category[, revenue := price * quantity]
  
  # Step 4: Aggregate total revenue by category
  revenue_by_category <- transactions_with_category[, .(total_revenue = round(sum(revenue, na.rm = TRUE), 2)), by = category]
  
  # Step 5: Sort by total revenue in decreasing order
  revenue_by_category <- revenue_by_category[order(-total_revenue)]
  
  return(revenue_by_category)
}

### Q4
# Function to calculate customers spending
calculate_spending_statistics <- function(transactions_data) {
  
  # Step 1: Filter customers with at least one transaction
  customers_with_transactions <- transactions_data[, .(total_spending = sum(price * quantity)), by = customer_id]
  
  # Step 2: Calculate summary statistics for monthly spending per customer
  average_spending <- mean(customers_with_transactions$total_spending)
  median_spending <- median(customers_with_transactions$total_spending)
  sd_spending <- sd(customers_with_transactions$total_spending)
  
  # Step 3: Return the results as a list
  results <- list(
    average_spending = round(average_spending, 2),
    median_spending = round(median_spending, 2),
    sd_spending = round(sd_spending, 2)
  )
  
  return(results)
}

### Q5
# Function to calculate revenue by gender
calculate_revenue_by_gender <- function(transactions_data, customers_data) {
  
  # Step 1: Merge transactions with customer data to get gender
  transactions_with_customers <- merge(transactions_data, customers_data[, .(customer_id, gender)], by = "customer_id", all.x = TRUE)
  
  # Step 2: Calculate the total revenue per transaction (price * quantity)
  transactions_with_customers[, revenue := price * quantity]
  
  # Step 3: Calculate total revenue by gender
  revenue_by_gender <- transactions_with_customers[, .(total_revenue = sum(revenue, na.rm = TRUE)), by = gender]
  
  # Step 4: Calculate total monthly revenue
  total_revenue <- sum(revenue_by_gender$total_revenue, na.rm = TRUE)
  
  # Step 5: Calculate percentage of total revenue for each gender
  revenue_by_gender$percentage <- round((revenue_by_gender$total_revenue/total_revenue)*100,2)
  
  revenue_by_gender$total_revenue <- round(revenue_by_gender$total_revenue,2)
  
  # Step 6: Return the revenue and percentage for each gender
  return(revenue_by_gender)
}

### Q6
# Function to calculate monthly spending per customer
calculate_monthly_spending_per_customer <- function(transactions_data, customers_data) {
  
  # Calculate total spending per customer (sum of price * quantity)
  transactions_data[, spending := price * quantity]
  monthly_spending <- transactions_data[, .(total_spending = sum(spending, na.rm = TRUE)), by = customer_id]
  
  # Merge with customers data to include demographic information (income, age, gender, region)
  customer_spending_data <- merge(monthly_spending, customers_data, by = "customer_id", all.x = TRUE)
  
  return(customer_spending_data)
}

# Function to fit a linear regression model
fit_spending_regression <- function(customer_spending_data) {
  # Scale income by 10,000 for easier interpretation
  customer_spending_data[, scaled_income := income / 10000]
  
  # Fit the linear regression model
  model <- lm(total_spending ~ scaled_income + age + gender + region, data = customer_spending_data)
  
  # Return the model summary
  return(summary(model))
}

### Q7
# Function to calculate total views, total purchases, and percentage of viewers who purchased
calculate_views_and_purchases <- function(views_data, transactions_data) {
  
  # Step 1: Identify customers with at least one view
  customers_with_views <- unique(views_data$customer_id)
  
  # Step 2: Identify customers with at least one purchase
  customers_with_purchases <- unique(transactions_data$customer_id)
  
  # Step 3: Calculate the number of customers who viewed at least 1 product
  total_customers_with_views <- length(customers_with_views)
  
  # Step 4: Identify customers who viewed a product and made a purchase
  customers_with_both <- intersect(customers_with_views, customers_with_purchases)
  total_customers_with_both <- length(customers_with_both)
  
  # Step 5: Calculate the percentage of customers who made a purchase among those with a view
  percentage_with_purchase <- round((total_customers_with_both / total_customers_with_views) * 100, 2)
  
  # Step 6: Return the results as a list
  result <- list(
    total_customers_with_views = total_customers_with_views,
    total_customers_with_purchases = total_customers_with_both,
    percentage_with_purchase = percentage_with_purchase
  )
  
  return(result)
}

### Q8
# Function to calculate viewing habits by age group
calculate_viewing_stats_by_age_group <- function(views_data, customers_data) {
  
  # Step 1: Create age group categories
  create_age_group <- function(age) {
    if (age >= 18 & age <= 34) {
      return("18-34")
    } else if (age >= 35 & age <= 49) {
      return("35-49")
    } else if (age >= 50 & age <= 64) {
      return("50-64")
    } else {
      return("65+")
    }
  }
  
  # Step 2: Merge views data with customers data to include age
  views_with_age <- merge(views_data, customers_data[, .(customer_id, age)], by = "customer_id", all.x = TRUE)
  
  # Step 3: Add age group column
  views_with_age[, age_group := sapply(age, create_age_group)]
  
  # Step 4: Calculate the number of unique products viewed per customer
  unique_products_per_customer <- views_with_age[, .(unique_products = uniqueN(product_id)), by = .(customer_id, age_group)]
  
  # Step 5: Calculate mean, median, and standard deviation for unique products viewed in each age group
  summary_stats <- unique_products_per_customer[, .(
    mean_unique_products = round(mean(unique_products),2),
    median_unique_products = round(median(unique_products),2),
    sd_unique_products = round(sd(unique_products),2)
  ), by = age_group][order(age_group)]
  
  # Step 6: Return the summary statistics
  return(summary_stats)
}

### Q9
# Function to calculate the increase in monthly average spending associated with an extra 10 product views
calculate_spending_increase_per_views <- function(views_data, transactions_data, customers_data) {
  
  # Step 1: Filter customers with at least 1 view
  customers_with_views <- unique(views_data$customer_id)
  
  # Step 2: Calculate total views per customer
  total_views_per_customer <- views_data[, .(total_views = .N), by = customer_id]
  
  # Step 3: Calculate monthly average spending per customer
  transactions_data[, spending := price * quantity]
  avg_spending_per_customer <- transactions_data[, .(total_spending = sum(spending, na.rm = TRUE), 
                                                     num_transactions = .N), by = customer_id]
  avg_spending_per_customer[, avg_spending := total_spending / num_transactions]
  
  # Step 4: Merge total views and average spending data with customer demographics
  customer_data <- merge(total_views_per_customer, customers_data, by = "customer_id", all.x = TRUE)
  customer_data <- merge(customer_data, avg_spending_per_customer[, .(customer_id, avg_spending)], by = "customer_id", all.x = TRUE)
  
  # Step 5: Handle missing (NA) values in the merged dataset
  customer_data[is.na(avg_spending), avg_spending := 0]   # Set avg_spending to 0 for customers with no purchases

  # Step 6: Fit the linear regression model
  model <- lm(avg_spending ~ total_views + age + gender + income + region, data = customer_data)
  
  # Step 7: Return the model summary
  return(summary(model))
}

### Q10
# Function to find customers who purchased the largest number of coats
find_top_coat_customers <- function(transactions_data, products_data) {
  
  # Step 1: Identify coat products in the products data
  coat_product_ids <- products_data[category == "coat", product_id]
  
  # Step 2: Join transactions with products to get the category information
  transactions_with_products <- merge(transactions_data, products_data[, .(product_id, category)], by = "product_id", all.x = TRUE)
  
  # Step 3: Filter for coat purchases
  coat_transactions <- transactions_with_products[category == "coat"]
  
  # Step 4: Group by customer and calculate the total number of coats purchased
  coats_per_customer <- coat_transactions[, .(total_coats_purchased = sum(quantity)), by = customer_id]
  
  # Step 5: Find the maximum number of coats purchased by any customer
  max_coats_purchased <- max(coats_per_customer$total_coats_purchased)
  
  # Step 6: Filter customers who purchased this maximum number of coats
  top_coat_customers <- coats_per_customer[total_coats_purchased == max_coats_purchased]
  
  # Step 7: Sort the result by customer_id in increasing order
  top_coat_customers <- top_coat_customers[order(customer_id)]
  
  # Step 8: Return the result
  return(top_coat_customers)
}

```

```{r load_data}
views <- fread(input = params$views_file)
transactions <- fread(input = params$transactions_file)
products <- fread(input = params$products_file)
customers <- fread(input = params$customers_file)
year <- params$year
month <- params$month
```

```{r explore_data, eval = FALSE}

```


```{r clean_data}
views[, time := as.POSIXct(time, format = "%Y-%m-%dT%H:%M:%OS")]

```

## `r year` - `r month`

## Specific Questions {.tabset}

This part of the report will be directed to product managers throughout the client's company.  The idea is to give them the useful information they need to act on the specific questions they posed.  Plan your communication accordingly.

For this report, make sure to use all of the data that was provided to you for the month.

### Q1

#### Question

During the first 7 days of the month, what were the 10 most viewed products?  Show the results in a table with the product's identifier, category, and count of the number of views.

#### Answer

```{r part2_q1}
first_7_top_10 <- get_top_10_views(views, products, year, month)
colnames(first_7_top_10)[colnames(first_7_top_10) == "product_id"] <- "Product ID"
colnames(first_7_top_10)[colnames(first_7_top_10) == "category"] <- "Category"
colnames(first_7_top_10)[colnames(first_7_top_10) == "view_count"] <- "View Count"
datatable(first_7_top_10)
```

### Q2

#### Question

During the whole month, what were the 10 most viewed products for each category?  Show the results in separate tables by category.  Include only the product's identifier and the count of the number of views.


#### Answer {.tabset}

```{r part2_q2_overall_calculations}
# Get the top 10 most viewed products for each category
top_10_by_category <- get_top_10_by_category(views, products)
```

##### Shirt

```{r part2_q2_shirt}
df_shirt = top_10_by_category$shirt
colnames(df_shirt)[colnames(df_shirt) == "product_id"] <- "Product ID"
colnames(df_shirt)[colnames(df_shirt) == "view_count"] <- "View Count"
datatable(df_shirt)
```

##### Pants

```{r part2_q2_pants}
df_pants = top_10_by_category$pants
colnames(df_pants)[colnames(df_pants) == "product_id"] <- "Product ID"
colnames(df_pants)[colnames(df_pants) == "view_count"] <- "View Count"
datatable(df_pants)
```

##### Coat

```{r part2_q2_coat}
df_coat = top_10_by_category$coat
colnames(df_coat)[colnames(df_coat) == "product_id"] <- "Product ID"
colnames(df_coat)[colnames(df_coat) == "view_count"] <- "View Count"
datatable(df_coat)
```

##### Shoes

```{r part2_q2_shoes}
df_shoes = top_10_by_category$shoes
colnames(df_shoes)[colnames(df_shoes) == "product_id"] <- "Product ID"
colnames(df_shoes)[colnames(df_shoes) == "view_count"] <- "View Count"
datatable(df_shoes)
```

##### Hat

```{r part2_q2_hat}
df_hat = top_10_by_category$hat
colnames(df_hat)[colnames(df_hat) == "product_id"] <- "Product ID"
colnames(df_hat)[colnames(df_hat) == "view_count"] <- "View Count"
datatable(df_hat)
```


### Q3

#### Question

What was the total revenue for each category during the month?  Show the results in a single table sorted in decreasing order.

#### Answer

```{r part2_q3}
revenue_by_category <- calculate_revenue_by_category(transactions, products)
# Display the results
revenue_by_category$total_revenue <- paste0("$", formatC(revenue_by_category$total_revenue, format = "f", big.mark = ",", digits = 2))
colnames(revenue_by_category)[colnames(revenue_by_category) == "category"] <- "Category"
colnames(revenue_by_category)[colnames(revenue_by_category) == "total_revenue"] <- "Total Revenue"
datatable(revenue_by_category)
```

### Q4

#### Question

Among customers with at least one transaction, show the average, median, and standard deviation of monthly spending per customer.

#### Answer

```{r part2_q4}
results <- calculate_spending_statistics(transactions)

# Display the results
# Create a data frame for the results
results_df <- data.frame(
  Metric = c("Average monthly spending per customer", 
             "Median monthly spending per customer", 
             "Standard deviation of monthly spending per customer"),
  Value = c(paste0("$",results$average_spending), 
            paste0("$",results$median_spending), 
            results$sd_spending)
)

datatable(results_df, rownames = FALSE, options = list(pageLength = 3, dom = 't', ordering = FALSE))

```

### Q5

#### Question

What percent of the monthly revenue was attributed to each category of gender?  Show the amount of revenue and the percentage for each group.

#### Answer

```{r part2_q5}
revenue_by_gender<- calculate_revenue_by_gender(transactions, customers)
revenue_by_gender$total_revenue <- paste0("$", formatC(revenue_by_gender$total_revenue, format = "f", big.mark = ",", digits = 2))
revenue_by_gender$percentage <- paste0(revenue_by_gender$percentage, "%")
colnames(revenue_by_gender)[colnames(revenue_by_gender) == "gender"] <- "Gender"
colnames(revenue_by_gender)[colnames(revenue_by_gender) == "total_revenue"] <- "Total Revenue"
colnames(revenue_by_gender)[colnames(revenue_by_gender) == "percentage"] <- "Percentage"
datatable(revenue_by_gender)
```

### Q6

#### Question

Using linear regression, what is the increase in monthly average spending associated with an extra ten thousand dollars of income?  Make sure your model also includes age, gender, and region as inputs to adjust for.

#### Answer

```{r part2_q6, echo=FALSE, results='asis'}
# Step 1: Calculate monthly spending data for customers in January 2020
customer_spending <- calculate_monthly_spending_per_customer(transactions, customers)

# Step 2: Fit the linear regression model and check the summary
regression_summary <- fit_spending_regression(customer_spending)
coefficients <- coef(regression_summary)
income_coef <- round(coefficients["scaled_income","Estimate"],2)

# Display the model summary
cat(sprintf("From the Estimate, increase in monthly average spending for every $10,000 increase in income is $ **%.2f**", income_coef))
df_summary <- round(regression_summary$coefficients[, c("Estimate", "Pr(>|t|)")],2)
colnames(df_summary)[colnames(df_summary) == "Pr(>|t|)"] <- "p-value"
datatable(df_summary)
cat(" Variables with p-value < 0.05 = statistically significant, it likely has a real effect on monthly average spending.")
```


### Q7

#### Question

Among customers who viewed at least 1 product, how many had at least one purchase during the month?  Show the total number of customers with a view, the total who made a purchase, and the percentage of customers with a view who made a purchase.

#### Answer

```{r part2_q7}
# Call the function
results <- calculate_views_and_purchases(views, transactions)

# Display the results
# Create a data frame for the results
results_df <- data.frame(
  Metric = c(
    "Total customers with at least 1 view", 
    "Total customers with at least 1 purchase", 
    "Percentage of customers with a view who made a purchase"
  ),
  
  Value = c(
    formatC(results$total_customers_with_views, format = "d", big.mark = ","), 
    formatC(results$total_customers_with_purchases, format = "d", big.mark = ","), 
    paste0(results$percentage_with_purchase, "%")
  )
)

datatable(results_df, rownames = FALSE, options = list(pageLength = 3, dom = 't', ordering = FALSE))
```

### Q8

#### Question

Now let's look at the viewing habits in different age groups, including 18-34, 35-49, 50-64, and 65+.  Within each group, what were the mean, median, and standard deviation for the number of unique products viewed per customer?

#### Answer

```{r part2_q8}
# Call the function and store the results
viewing_stats_by_age_group <- calculate_viewing_stats_by_age_group(views, customers)

# Display the results
colnames(viewing_stats_by_age_group)[colnames(viewing_stats_by_age_group) == "age_group"] <- "Age Group"
colnames(viewing_stats_by_age_group)[colnames(viewing_stats_by_age_group) == "mean_unique_products"] <- "Mean Unique Products"
colnames(viewing_stats_by_age_group)[colnames(viewing_stats_by_age_group) == "median_unique_products"] <- "Median Unique Products"
colnames(viewing_stats_by_age_group)[colnames(viewing_stats_by_age_group) == "sd_unique_products"] <- "SD Unique Products"

datatable(viewing_stats_by_age_group)

```


### Q9 

#### Question

Using linear regression, what is the increase in a customer's monthly average spending associated with an extra 10 total product views?  Make sure your model also includes age, gender, income, and region as inputs to adjust for.  Use only customers with at least 1 view during the month.  Use a spending value of 0 for customers without transactions during the month.  Make sure to round your answer to 2 decimal places.

#### Answer

```{r part2_q9, echo=FALSE, results='asis'}
# Call the function and store the result
spending_increase_per_10_views <- calculate_spending_increase_per_views(views, transactions, customers)

# Interpret the coefficient for total views
coefficients <- coef(spending_increase_per_10_views)
views_coef <- coefficients["total_views","Estimate"]

# Multiply by 10 to get the effect of 10 additional product views
total_views_coef_10 <- round(views_coef * 10,2)

# Display the model summary
cat(sprintf("From the Estimate, a customer's monthly average spending associated with an extra 10 total product views is $ **%.2f**", total_views_coef_10))

df_summary <- round(spending_increase_per_10_views$coefficients[, c("Estimate", "Pr(>|t|)")],3)
colnames(df_summary)[colnames(df_summary) == "Pr(>|t|)"] <- "p-value"
datatable(df_summary)
cat(" Variables with p-value < 0.05 = statistically significant, it likely has a real effect on monthly average spending.")
```

### Q10

#### Question

Which customer(s) purchased the largest number of coats during the month?  In the event of a tie, include all of the users who reached this value, listed in increasing sorted order.  Show their identifiers and total volume of coats purchased during the month.

#### Answer

```{r part2_q10}
# Call the function and store the result
top_customers <- find_top_coat_customers(transactions, products)

# Display the result

colnames(top_customers)[colnames(top_customers) == "customer_id"] <- "Customer ID"
colnames(top_customers)[colnames(top_customers) == "total_coats_purchased"] <- "Total Coats Purchased"

datatable(top_customers)
```

